Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-18 docs/proposals/NSF/ProjDesc/ProjDesc.tex
NM

In particular, we view the slow, accumulated effects of erosion and dissolution as instrumental in altering porous-medium properties. First, by deteriorating solid material, erosion alters the size distribution of grains, enabling smaller grains to be sieved through networks of larger ones. A second and more subtle way that erosion can alter medium properties is by changing the {\em shape} distribution of grains. For instance, erosion can carve grains into slender shapes aligned primarily in the direction of the dominant flow, thereby creating {\em anisotropic} characteristics. Likewise, chemical dissolution can carve well-defined conduit paths in karst networks which strongly alter transport properties~\cite{szymczak2009wormhole}. Some visible effects of flow-induced erosion are shown in Fig.~\ref{collage}. The top row shows products of primarily {\em chemically-driven} erosion, while the bottom shows effects of primarily {\em mechanically-driven} erosion. Throughout this proposal, we also refer to chemically-driven erosion as dissolution. Some examples in the figure include the rough surface of limestone pavement, a sinkhole, an erosion-curved conduit, and finally an image of the small-scale structure of soil, exhibiting polydispersity and anisotropy, resulting from a combination of erosive, compaction, and transport properties

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/collage.pdf}
\caption{\label{collage} Geophysical examples of chemical and mechanical erosion. Top from left: Diagram of a coupled karst-soil system with conduit formation \cite{watson2019sinkholes}; Limestone pavement in Dent de Crolles, France; Nohoch Ch’en Sinkhole in Belize. Bottom: examples of soil erosion. The bottom right shows the microscopic soil texture.}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

\paragraph{Relevance to the Computational Mathematics Synopsis:}
\begin{itemize}[noitemsep]
  \item Efficient, high-fidelity computational methods form the core of the proposed work of porous-media erosion. Large ensembles of the high-fidelity simulations will inform the coarse-grained models that will be developed in the later stages of the proposed work.
  \item The proposed research is interdisciplinary by nature, with immediate applications in the fields of geophysics and environmental sciences. The PIs already have experience collaborating with scientists from other disciplines through FSU's Geophysical Fluid Dynamics Institute, an interdisciplinary unit comprised of a diverse set of faculty from home departments in example Earth, Atmospheric, and Ocean Sciences, Geology, Physics, Civil Engineering, Environmental Engineering, and others.
  \item The proposed research is a collaborative effort with a focus on training future generations of scientists and mathematicians. The proposal will include graduate, undergraduate, and K--12 students, with close cooperation between them.
\end{itemize}

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/80circ8vort.pdf}
\caption{\label{80circ8vort} Simulation of 80 bodies eroding in Stokes flow under the action of shear stress. Flow is left to right. Color represents vorticity, which provides a convenient way to visualize local shear rates. 
}
\end{center}
\end{figure}
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/fig3.pdf}
  \caption{\label{fig3} {\em Left}: The area fraction of solid bodies versus time. Time is normalized by the vanishing time $t_f$. {\em Middle}: The resistivity of the porous matrix decrease much more rapidly than area fraction, indicating the reshaping process to be essential. Right: The anisotropy (ratio of vertical to horizontal resistivity) as it varies in time. Erosion can create a highly anisotropic medium.}
\end{center}
\end{figure}
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

%----------------------------------------------------------------------%
\subsection{Transport through eroded media}

@@

\todo[inline]{Bryan, do you know any other references to include after `topic of current research'? I have Shelley and Yan's method. There were a bunch of others at Oaxaca but I don't remember names.}

@@

%----------------------------------------------------------------------%
\subsection{Undergraduate Research}
Regarding undergraduate research, the ideas in this proposal represent a rich and fascinating topic for undergraduate students, while also being approachable these students. In particular, we aim to recruit at least one undergraduate student to perform laboratory experiments at the Geophysical Fluid Dynamics Institute (GFDI). As mentioned above, GFDI offers state-of-the-art laboratory facilities, with an already-functioning sinkhole experiment. The student will perform controlled experiments using natural and artificial porous media (e.g.~sand and micro-glass beads respectively) to precisely characterize the parameter regime leading to gravitational collapse. Parallel experiments will be conducted with erodible (bentonite clay) or dissolvable  (sucrose solids) fragments materials, which will represent the porous medium. These experiments will be used to guide, calibrate, and validate the computational advances. These undergraduate students will work closely with the graduate students to ensure that theory and experiments advance in close cooperation, with constant feedback between the two.

>>>>

In particular, we view the slow, accumulated effects of erosion and dissolution as instrumental in altering porous-medium properties. First, by deteriorating solid material, erosion alters the size distribution of grains, enabling smaller grains to be sieved through networks of larger ones. A second and more subtle way that erosion can alter medium properties is by changing the {\em shape} distribution of grains. For instance, erosion can carve grains into slender shapes aligned primarily in the direction of the dominant flow, thereby creating {\em anisotropic} characteristics. Likewise, chemical dissolution can carve well-defined conduit paths in karst networks which strongly alter transport properties~\cite{szymczak2009wormhole}. Some visible effects of flow-induced erosion are shown in Fig.~\ref{collage}. The top row shows products of primarily {\em chemically-driven} erosion, while the bottom shows effects of primarily {\em mechanically-driven} erosion. Throughout this proposal, we also refer to chemically-driven erosion as dissolution. Some examples in the figure include the rough surface of limestone pavement, a sinkhole, an erosion-curved conduit, and finally an image of the small-scale structure of soil, exhibiting polydispersity and anisotropy that result from a combination of erosive, compaction, and transport actions.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/collage.pdf}
\caption{\label{collage} Geophysical examples of chemical and mechanical erosion. Top from left: Diagram of a coupled karst-soil system with conduit formation \cite{watson2019sinkholes}; Limestone pavement in Dent de Crolles, France; Nohoch Ch’en Sinkhole in Belize. Bottom: examples of soil erosion. The bottom right shows the microscopic soil texture, exhibiting polydispersity and anisotropy.}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

\paragraph{Relevance to the Computational Mathematics Synopsis:}
\begin{itemize}[noitemsep]
  \item Efficient, high-fidelity computational methods form the core of the proposed work of porous-media erosion. Large ensembles of the high-fidelity simulations will inform the coarse-grained models that will be developed in the later stages of the proposed work.
  \item The proposed research is interdisciplinary by nature, with immediate applications in the fields of geophysics and environmental sciences. The PIs already have experience collaborating with scientists from other disciplines \cite{Caves2018, eastham2019multiphase, Bolles2019} through FSU's Geophysical Fluid Dynamics Institute, an interdisciplinary unit comprised of a diverse set of faculty from home departments in example Earth, Atmospheric, and Ocean Sciences, Geology, Physics, Civil Engineering, Environmental Engineering, Scientific Computing, Mathematics, and others.
  \item The proposed research is a collaborative effort with a focus on training future generations of scientists and mathematicians. The proposal will include graduate, undergraduate, and K--12 students, with close cooperation between them.
\end{itemize}

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/80circ8vort.pdf}
\caption{\label{80circ8vort} Simulation of 80 bodies eroding in Stokes flow under the action of shear stress. Flow is left to right. Color represents vorticity, which provides a convenient way to visualize local shear rates. 
}
\end{center}
\end{figure}
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htp]
\begin{center}
\includegraphics[width = 0.99 \textwidth]{./figs/fig3.pdf}
  \caption{\label{fig3} {\em Left}: The area fraction of solid bodies versus time. Time is normalized by the vanishing time $t_f$. {\em Middle}: The resistivity of the porous matrix decrease much more rapidly than area fraction, indicating the reshaping process to be essential. Right: The anisotropy (ratio of vertical to horizontal resistivity) as it varies in time. Erosion can create a highly anisotropic medium.}
\end{center}
\end{figure}
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

%----------------------------------------------------------------------%
\subsection{Dispersive transport through eroded media}

@@

\todo[inline]{Bryan, do you know any other references to include after `topic of current research'? I included Shelley and Yan's method. There were a bunch of others at Oaxaca but I don't remember names.}

@@

%----------------------------------------------------------------------%
\subsection{Undergraduate Research}
Regarding undergraduate research, the ideas in this proposal represent a rich and fascinating topic for undergraduate students, while also being approachable by the skills acquired at that stage. In particular, we aim to recruit at least one undergraduate student to perform laboratory experiments at the Geophysical Fluid Dynamics Institute (GFDI). As mentioned above, GFDI offers state-of-the-art laboratory facilities, with an already-functioning sinkhole experiment. The student will perform controlled experiments using natural and artificial porous media (e.g.~sand and micro-glass beads respectively) to precisely characterize the parameter regime leading to gravitational collapse. Parallel experiments will be conducted with erodible (bentonite clay) or dissolvable  (sucrose solids) fragments materials to represent a {\em dynamically} changing porous medium. These experiments will be used to guide, calibrate, and validate the computational advances. These undergraduate students will work closely with the graduate students to ensure that theory and experiments advance in close cooperation, with constant feedback between the two.
